Trygve's Problem

### 2018-03-12 trygve/outline.tex
trygve

\usepackage{amsthm}
\newtheorem{thm}{Theorem}

@@

When many different measurement tools are available, it can be difficult to understand
how to combine information from different tools.  This challenge is compounded when the measurement
tools are destructive.  For example, after processing a neuron with a given single-cell rna
sequencing technique, we cannot generally go back and process that same neuron with a different
technique.  This is because the measurement technique destroys the original neuron.  This means
we can never see \emph{both measurements} for the same cell, side by side.  How, then, can we ever
hope to understand how the different measurements relate to each other?

In this paper, we hope to shed some light on this this basic mathematical question.  It has many
applications in the hard sciences.  For example:

@@

How could we learn the relationship between two measurement techniques, unless we can observe both measurements 
applied to the same members of a population (dual measurement)?  In this work we show that it is indeed 
sometimes possible to get useful and statistically rigorous bounds, even without these dual measurements.

@@

We then apply our method to single cell RNA data.  We show how different clusterings from different single 
cell RNA modalities can be connected.  Our method yields results which are generally consistent with the scientist's
understanding of the data and clusters, but also reveal potential gaps which could be important for further study.

@@

We can now articulate our goal.  Informally, our goal is to see what we can understand about the relationship between
the observations $X_i$ from one measurement modality and the other measurement modality $Y_i$.  Ideally, we would
like to learn the distribution $p^*(y_i|x_i)$.  Unfortunately, given the limitations of the data, this may not be 
possible.  Instead, we articulate a slightly less amibtious goal: to put \emph{bounds} on $q(y_i|x_i)$.  Specifically,
we will seek to produce functions $\hat u_1(y|x),\hat u_2(y|x)$ such that

@@

In particular, we define

@@

These will serve as consistent estimators for our quantities of interest, $u_1^*,u_2^*$, as the following theorem shows.\vspace{.1in}

\begin{thm}
Fix any $\kappa>0$.  Let $N_{X,\ell},N_{Y,\ell}\rightarrow\infty$ uniformly in $\ell$.  Then $\hat u_1(y|x)\rightarrow u^*_1(y|x)$ in probability, and likewise for $u_2$.
\end{thm}

\begin{proof}

The logic is in perfect symmetry for $u_1,u_2$, so we focus on the $u_1$ case.

We begin by seeing how our process would have worked if we were able to know the exact values of $p^*$ and $h^*$,
instead of the empirical distributions $\hat p,\hat h$.  That is, we define
%
\begin{align*}
L_N(q,p,h) &\triangleq \sum_\ell N_{Y,\ell}\sum_{y}h(y|\ell)\log\left(\sum_{x}p(x|\ell)q(y|x)\right) + \kappa \sum_{xy} \log q(y|x) \\
\Theta(p,\tilde q) &\triangleq\left\{q:\ \sum_x p(x|\ell) \tilde q(y|x) =\sum_x p(x|\ell) q(y|x), \forall \ell,y\right\}
\end{align*}
%
and take
%
\begin{align*}
\tilde q^* &= \argmax_q L_N(q,p^*,h^*)  \\
\tilde u_1^*(y|x) &= \inf_{q\in \Theta(p^*,\tilde q^*)} q(y|x) \\
\end{align*}
%
Our first task is to show that the map $p^*,h^* \mapsto \tilde u_1^*$ is continuous.  This follows in two steps:

%%%%% two steps
\begin{enumerate}
% STEP I
\item The map $p^*,h^* \mapsto \argmax_q L(q,p^*,h^*)$ is continuous.  Note that $q\mapsto L(q,p^*,h^*)$
is \emph{uniformly} strictly concave for all $p^*,h^*$.  Indeed, the hessian is given by
%
\begin{align*}
\frac{\partial L_N(q,p^*,h^*)}{\partial q(x_1,y_1) \partial q(x_2,y_2)} =& 
    - \delta_{y_1=y_2}\sum_\ell N_{Y,\ell} h^*(y_1|\ell) 
              \frac{p^*(x_1|\ell)p^*(x_2|\ell)}{\left(\sum_x p(x|\ell) q(y_1|x)\right)^2} \\
    & \qquad - 2\kappa \delta_{(x_1,y_1)=(x_2,y_2)} \\
\end{align*}
%
Since $q$ lives in a compact space and $L$ is continuous with respect to $p^*,h^*$, we thus 
know that $p^*,h^* \mapsto \argmax_q L_N(q,p^*,h^*)$ is also continuous.  
% STEP II
\item The map $p,\tilde q \mapsto \inf_{q\in\Theta(p,\tilde q)} q(y|x)$ is continuous, as long as 
$\{(p(x_1|\ell)\cdots p(x_n|\ell))\}_\ell$ are linearly independent (remember we have assumed the independence). 
It would suffice to show the following: 
for any fixed $p_1 \approx p_2,\tilde q_1 \approx \tilde q_2,x,y$ where the $\approx$ is sufficiently tight
and $q_1\in \Theta(p_1,\tilde q_1)$, 
we can find $q_2\in \Theta(p_2,\tilde q_2)$ with $q_1(y|x)\approx q_2(y|x)$.  We will solve
this problem first by looking at what happens if $\tilde q$ changes, and then at what happens if $p$ changes.
\begin{itemize}
\item Let us say $p_1=p_2=p$.  Then, in the absence of constraints, we could solve
our problem by simply taking $q_2 = q_1 + \tilde q_2 - \tilde q_1$.
However, if $q_1(y|x)=0$ for any $x,y$, we cannot guarantee that the resulting $q_2$ will be a positive measure. 
To resolve this, we note that the entropic penalty ensures that $\tilde q_2,\tilde q_1$ lie on the interior
of the probability simplex.  This enables us to resolve the issue by constructing ``smoothed'' version of $q_1$, namely
$(1-\alpha) q_1 + \alpha (\tilde q_1)$ which also lies on the interior of the probability simplex.   Let's do this
rigorously.

Fix any $\tilde q_1,\delta>0$ and $q_1 \in \Theta(p,\tilde q_1)$. Let $c = \min_{x,y} \tilde q_1(y|x)$.  Now pick any $\tilde q_2$ such that $|\tilde q_2-\tilde q_1|< \delta c/2$ in the uniform norm.  Take $q_2 = (1-\delta)q_1 + \delta \tilde q_1 + \tilde q_2 - \tilde q_1$.  Observe that $q_2 (y|x) \geq \delta c - \delta c/2$ for every $x,y$, i.e. it is a valid
probability density.  Note also that $q_2 \in \Theta(p, \tilde q_2)$.  Finally, note that $|q_2 - q_1|<\delta$ in
the uniform norm, as desired.

\item Let us now say $\tilde q_1 = \tilde q_2 = \tilde q$.  We are given $q_1,p_1,p_2$ and asked to
produce $q_2\in \Theta(p_2,\tilde q)$ such that $q_1\approx q_2$.  Let 
$A_{1,\ell,x} = p_1^*(x|\ell)$ and $A_{2,\ell,x} = p_2^*(x|\ell)$.  We have assumed that the rows of $A_1$ are linearly independent, so $A_1 A_1^T$ is invertible.  There is therefore an open set around $A_1$ so that as long as $A_2 \approx A_1$ we have that the rows of $A_2$ are also linearly independent.  Therefore to find $q_2 \in \Theta(p_2, \tilde q)$
we can simply take $q_2 = q_1 - A_2^T (A_2 A_2^T)^{-1} A_2 (q_1 -\tilde q)$.  Note that $A_1 (q_1 - \tilde q) =0$, so
the norm of $A_2 (q_1 -\tilde q)$ can be bounded.  Likewise the strength of $(A_2 A_2^T)^{-1}$
can be bounded by the smallest absolute eigenvalue of $A_2$, which, in turn, is bounded by its proximity to $A_1$.  
Finally, the entries of $A_2^T$ are uniformly bounded between zero and one.  In conclusion, 
for any $p_1$ we can always find a $p_2$ close enough so that we can completely control $|q_1-q_2|$ in the Euclidean
norm.  Since we are working in finite dimensions, this is sufficient to control it in the uniform norm.

Finally, we note that it is possible this method will result in $q_2$ being an invalid probability distribution, i.e.
$q_2(y|x)<0$.  This could be true in two senses.  First, it might be that $\sum_y q_2(y|x)\neq 1$.  Fortunately,
this is impossible.  Indeed, $(q_1 - \tilde q)1 =0$ so certainly $(q_2-q_1)1=(\cdots)(q_1-\tilde q)1=0$ as well.   
Second, we might have that $q_2(y|x)<0$.  To remedy this
second point, we take recourse to the same method explained in detail for the case $p_1=p_2$ above. 
That is, we work with a slightly ``smoothed'' version of $q_1$, and find something which is very close to the smoothed version,
and thereby ensure that the resulting $q_2$ is valid.  This $q_2$ is close to the slightly smoothed $q_1$ and thus also
close to $q_1$.

\end{enumerate}

It is well-known that the empirical distributions $\hat p, \hat h$ converge uniformly in probability to the true values, $p^*,h^*$.  Since the map $p,q \rightarrow u_1$ is continuous in a neighborhood of $p^*,q^*$, it follows that $\hat u_1$ also converge uniformly to $\tilde u_1^*$ in probability.

It thus suffices to show that $\tilde u_1^* \rightarrow u_1^*$.  To show this, let $\tilde h^* =p^*\tilde q^*$,
i.e. $h^*(y|\ell)=\sum_x p^*(x|\ell)\tilde q^*(y|x)$.  The key will be to control $\tilde h^*$ and see how that control
enables us to control $\tilde u^*$.

\begin{enumerate}
\item $\tilde h^* \rightarrow h^*$ in the uniform norm as 
$N_{Y,\ell}\rightarrow\infty$ uniformly in $\ell$.

To show this, we express $L$ in terms of 
Kullback-Leibler divergences, $D(h^*_\ell||\tilde h^*_\ell)$ and entropies 
$\mathcal H(h^*_\ell)$.  Applying Pinsker's inequality, we obtain
%
\begin{align*}
L(\tilde q^*,p^*,h^*) &= \sum_\ell N_{y,\ell} (\mathcal H(h^*_\ell) - D(h^*_\ell||\tilde h^*_\ell)) + \kappa \sum_{xy} \log q^*(y|x)\\
&\leq \sum_\ell N_{y,\ell} (\mathcal H(h^*_\ell) - 2|h^*_\ell-\tilde h^*_\ell|^2_{TV} + \kappa \sum_{xy} \log q^*(y|x)\\
&\leq \sum_\ell N_{y,\ell} (\mathcal H(h^*_\ell) - \frac{1}{2}(\sup_y |h^*(y|\ell)-\tilde h^*(y|\ell)|)^2 + \kappa \sum_{xy} \log q^*(y|x)\\
\end{align*}

Since $\tilde q^*,\tilde h^*$ are found by taking $L$ as large as possible, it would thus seem that we could use
this inequality to argue that $h^* \approx \tilde h^*$ when $N_{Y,\ell}$ are large.  There is, however, once again 
a bit of trickiness in the case that $q^*(y|x)=0$ for some $x,y$.  Indeed, it may be that $L(q^*,p^*,h^*)=-\infty$,
and this complicates some of the arguments that we want to make.   To remedy this, we first design a smoothed
version of $q^*$.  Then everything goes through as we might like.  Let's do it rigorously.

Fix $0<\epsilon<1$. We take
$q^*_{SM}= (1-\epsilon) q^* + \epsilon \tilde q^*$ as our smoothed version of $q^*$.  
The convexity of the KL divergence then gives that
%
\begin{align*}
D(h^* || p^*q^*_{SM})&=D(h^*_{\ell} || (1-\epsilon) h^*_{\ell} + \epsilon \tilde h^*) \leq (1-\epsilon)0 + \epsilon D(h^*_{\ell} || \tilde h^*) 
\end{align*}
%
Thus, applying Pinsker (and the fact that $\tilde q^* = \argmax _q L_N(\tilde q,p^*,h^*)$)
%
\begin{align*}
0 &\leq L_N(\tilde q^*,p^*,h^*) - L_N(q^*_{SM},p^*,h^*)\\
 &= \sum_\ell N_{y,\ell} (\epsilon-1) D(h^*_\ell||\tilde h^*_\ell)+ \kappa \sum_{xy} \log \frac{\tilde q^*(y|x)}{q^*_{SM}(y|x)}\\
&\leq \sum_\ell N_{y,\ell} (\epsilon-1) D(h^*_\ell||\tilde h^*_\ell)- \kappa |\Omega_X||\Omega_Y|\log \epsilon \\
&\leq \sum_\ell N_{y,\ell} (\epsilon-1) 2|h^*_\ell-\tilde h^*_\ell|^2_{TV}- \kappa |\Omega_X||\Omega_Y|\log \epsilon \\
&\leq \frac{\epsilon-1}{2}\sum_\ell N_{y,\ell} \left(\sup_y|h^*(y|\ell)-\tilde h^*(y|\ell)|\right)^2-\kappa |\Omega_X||\Omega_Y|\log \epsilon \\
\end{align*}
%
One consequence of that inequality is that 
%
\begin{align*}
\sup_y|h^*(y|\ell)-\tilde h^*(y|\ell)| &\leq \sqrt{\frac{-2\kappa |\Omega_X||\Omega_Y|\log \epsilon}{N_{Y,\ell}} } \\
\end{align*}
%
With this inequality in hand, we are now prepard to prove that $\tilde h^* \rightarrow h^*$.  Pick any $0<\delta<1$.  
Choose $N_{\ell,y}>-2\kappa |\Omega_X||\Omega_Y|\log \delta/\delta$.  Take $\epsilon=\delta$.  Then
%
\begin{align*}
\sup_y|h^*(y|\ell)-\tilde h^*(y|\ell)| &\leq \sqrt{\delta} \\
\end{align*}

\item We now show that if $\tilde h^*\approx h^*$, then $\tilde u_1^* \approx u_1^*$.  To do so, let
%
\[
\tilde \Theta^* = \left\{q: \sum_x p^*(x|\ell)q(y|x)=\sum_x p^*(x|\ell)\tilde q^*(y|x) = \tilde h^*(y|\ell)\right\} 
\]
%
To complete our proof, it then suffices to show that we can find $\tilde h^*$ close enough to $h^*$ so that
we can find a $q \in \tilde \Theta^*$ with $q \approx q^*$.

\end{enumerate}

\end{proof}

Then certainly $u_1^*(y|x) \leq q(y|x)\leq u_2^*(y|x)$.  What's more, we can

>>>>

\usepackage{amsthm}
\newtheorem{thm}{Theorem}
\newtheorem{conj}{Conjecture}

@@

When many different measurement tools are applied, how can we best combine information from the different different tools?  If we can apply various measurement tools to the same specimens, we can at least begin to understand how the tools they are related.  We call this ``joint measurement.''  However, the challenge is compounded when joint measurement is unavailable.  For example, consider the case that the measurement tools are destructive.  After processing a neuron with a given single-cell rna sequencing technique, we cannot generally go back and process that same neuron with a different technique.  This is because the measurement technique destroys the original neuron.  This means we can never see \emph{both measurements} for the same cell, side by side.  How, then, can we ever hope to understand how the different measurements relate to each other?

In this paper, we hope to shed some light on this this basic mathematical question.  It has many applications in the sciences.  For example:

@@

In this work we show that it is indeed sometimes possible to get useful and statistically rigorous bounds for these kinds of questions, even when joint measurements are unavailable.

@@

We apply our method to single cell RNA data.  We show how different clusterings from different single cell RNA modalities can be connected.  Our method yields results which are generally consistent with the scientist's understanding of the data and clusters, but also reveal potential gaps which could be important for further study.

@@

We can now articulate our goal.  Informally, our goal is to see what we can understand about the relationship between the observations $X_i$ from one measurement modality and the other measurement modality $Y_i$.  Ideally, we would like to learn the distribution $p^*(y_i|x_i)$.  Unfortunately, given the limitations of the data, this may not be possible.  Instead, we articulate a slightly less amibtious goal: to put \emph{bounds} on $q(y_i|x_i)$.  Specifically, we will seek to produce $\hat u_1(y|x),\hat u_2(y|x)$ such that

@@

with high probability.  In particular, we define

@@

We conjecture that these will serve as consistent estimators for our quantities of interest, $u_1^*,u_2^*$.\vspace{.1in}

\vspace{.1in}
\begin{conj}
Fix any $\kappa>0$.  Let $N_{X,\ell},N_{Y,\ell}\rightarrow\infty$ uniformly in $\ell$.  Then $\hat u_1(y|x)\rightarrow u^*_1(y|x)$ in probability, and likewise for $\hat u_2,u^*_2$.
\end{conj}

\section{Empirical results}

...look reasonable, and more or less checked out with the Allen Institute people.  That said, they caution that some cell types tend to die more often in one experimental modality than another, which would violate our assumptions.  However, this doesn't bother them, because they see this technique as a way to \emph{validate} their assignments of clusters.  As such, if the result of our algorithm doesn't make sense, that's great because it may help them to figure out \emph{which cell types are dying more} and then they can upweight certain groups correspondingly.  Or it may help them figure out that their cluster assignment was all wrong.  This is how they see my method fitting into their pipelines: as a way of sanity checking their cluster assignments.  Which seems cool.

\section{The conjecture}

Here we give an outline for how the conjecture might be proved.

For any transition matrices $p,h$,\footnote{Here by transition matrix we just mean that the rows sum to 1 and the entries are nonnegative.} let $\Theta(p,h)$ denote the subset of transition matrices satisfying

\[
\Theta(p,h) \triangleq \{q:\ pq=h\}
\]

Now let us say we have some $p^*,h^*$, and some estimators $\hat p,\hat h$.  If the estimators were consistent, we might hope that $\Theta(\hat p,\hat h)$ would yield a consistent estimator for $\Theta(p^*,h^*)$.  In particular, we would ask that the Hausdorff distance would converge to zero in probability, $d_H(\Theta(\hat p,\hat h),\Theta(p^*,h^*))\rightarrow 0$.  Unfortunately, this is simply not the case.  Here's an example:

\begin{align*}
p^* &= (1, 0)\\
h^* &= (1, 0)\\
\Theta(p^*,h^*) &= 
  \left\{q:\ q=\left(\begin{array}{cc}
  1 & 0 \\
  \alpha & 1-\alpha \\
  \end{array}\right)\right\}
\end{align*}

Notice that the second row of $q$ is left \emph{totally} unspecified.  This is because the equation $p^*q=h^*$ simply places no constraints on the second row of $q$.  Now let us say we use our consistent estimator, and get something \emph{very} close to the truth.  We might get a dreadfully wrong impression of $q$:

\begin{align*}
\hat p &= \left(\frac{99}{100}, \frac{1}{100}\right)\\
\hat h &= (1, 0)\\
\Theta(\hat p,\hat h) &= 
  \left\{q:\ q=\left(\begin{array}{cc}
  1 & 0 \\
  1 & 0 \\
  \end{array}\right)\right\}
\end{align*}

Suddenly $q$ is \emph{entirely constrained} by the equation $pq=h$.  This dramatic change is possible because $q$ is constrained not only by the equation $pq=h$ but also by the fact that $q$ must be a transition matrix.  This allows for the possibility that although $p^*\approx \hat p$ and $h^*=\hat h$, we have that $d_H(\Theta(\hat p,\hat h),\Theta(p^*,h^*))=1$.  Not good.

However, this troubling example relied crucially on the fact that a given entry of $p$ went from zero to nonzero.  Let us therefore ask that the estimator $\hat p$ was consistent both in the usual Euclidean $\mathscr L^2$ sense, as well as in the sense that it eventually got the zeros right.  That is, $\mathbb P(\hat p(x|\ell)=0\Leftrightarrow p^*(x|\ell)=0) \rightarrow 1$.  Such an estimator is certainly easy to come by (the empirical distribution will do).  To prove consistency, it would then be sufficient to show that

\vspace{.1in}
\begin{conj}
For every $\epsilon>0$, we can find a $\delta>0$ so that if $|h_1 -h_2|^2<\delta$, $|p_1 -p_2|^2<\delta$, and $p_1(x|\ell)=0\Leftrightarrow p_2(x|\ell)=0$, then $d_H(\Theta(p_1,h_1),\Theta(p_1,h_1))<\epsilon$.
\end{conj}

To prove this conjecture, the essential point is to understand how the constraint $pq=h$ related to the $k$-faces (vertices, facets, etc) of the polytope of transition matrices.  If that constraint is somehow \emph{degenerate} with respect to one of the faces, then we get in trouble because a slight perturbation can change the degeneracy and thus create weird effects.  However, as long as the nature of the degeneracy is the same for both $p_1,p_2$, I believe there is no problem.  So to prove this conjecture, it remains to do three essential things

\begin{itemize}
    \item mathematically characterize the ``degeneracy'' of the constraint $pq=h$ with respect to faces of the transition matrix polytope
    \item show that if $p_1,p_2$ share the same qualitative nature of degeneracy and $p_1\approx p_2$, then $\Theta(p_1,h_1)\approx\Theta(p_2,h_2)$.
    \item show that the qualitative nature of this degeneracy is only changed when an entry of $p$ switches between zero and nonzero
\end{itemize}
